---
title: "An R Package for fast segmentation"
authors:
  - name: "Günter Klambauer"
    affiliation: "Institute of Bioinformatics, Johannes Kepler University Linz
                  Altenberger Str. 69, 4040 Linz, Austria"
  - name: "Andreas Mitterecker"
    affiliation: "Institute of Bioinformatics, Johannes Kepler University Linz
                  Altenberger Str. 69, 4040 Linz, Austria"
date: "`r format(Sys.time(), '%B %d , %Y')`"
package: fastseg
vignette: >
  %\VignetteIndexEntry{An R Package for fast segmentation}
  %\VignetteIndexEntry{fastseg: Manual for the R package}
  %\VignetteDepends{fastseg}
  %\VignettePackage{fastseg}
  %\VignetteKeywords{copy number analysis, segmentation,
  % CNV, copy number variant}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document
bibliography: fastseg.bib
reference-section-title: References
link-citations: yes
csl: bioinformatics.csl
---

## Scope and Purpose of this Document {#scope-and-purpose-of-this-document .unnumbered}

This document is a user manual for the R package `r Biocpkg("fastseg")`. It is
only meant as a gentle introduction into how to use the basic functions
implemented in this package. Not all features of the R package are described in
full detail. Such details can be obtained from the documentation enclosed in the
R package. Further note the following: (1) this is neither an introduction to
segmentation algorithms; (2) this is not an introduction to R. If you lack the
background for understanding this manual, you first have to read introductory
literature on these subjects.

# Introduction

`r Biocpkg("fastseg")` implements a very fast and efficient segmentation
algorithm. It has similar functionality as `r Biocpkg("DNACopy")` [@Olshen:04]
but is considerably faster and more flexible. `r Biocpkg("fastseg")` can segment
data stemming from DNA microarrays and data stemming from next generation
sequencing for example to detect copy number segments. Further it can segment
data stemming from RNA microarrays like tiling arrays to identify transcripts.
Most generally, it can segment data given as a matrix or as a vector. Various
data formats can be used as input to `r Biocpkg("fastseg")` like expression set
objects for microarrays or GRanges for sequencing data.

The segmentation criterion of `r Biocpkg("fastseg")` is based on a statistical
test in a Bayesian framework, namely the cyber t-test [@Baldi:01]. The speed-up
stems from the facts, that sampling is not necessary in for 
`r Biocpkg("fastseg")` and that a dynamic programming approach is used for
calculation of the segments' first and higher order moments.

For further information regarding the algorithm and its assessment see the
fastseg homepage at http://www.bioinf.jku.at/software/fastseg/fastseg.html

# Getting started

To load the package, enter the following in your R session: 

```{r message=FALSE}
library(fastseg)
```

## Data

According to the `r Biocpkg("DNAcopy")` package from bioconductor we selected a
subset of the data set presented in [@Snijders:01]. This data set will be called
`coriell`. The data correspond to two array CGH studies of fibroblast cell
strains. [^1] In particular, the studies **GM05296** and **GM13330** were
chosen. After selecting only the mapped data from chromosomes 1-22 and X, there
are 2271 data points.

To prepare the data for our examples we execute the following code:

```{r echo=TRUE}
data(coriell)
head(coriell)

samplenames <- colnames(coriell)[4:5]
data <- as.matrix(coriell[4:5])
#data[is.na(data)] <- median(data, na.rm=TRUE)
chrom <- coriell$Chromosome
maploc <- coriell$Position
```

The main functions of the package are `fastseg` and `toDNAcopyObj`. The first on
runs the segmentation algorithm and the latter converts the segmentation results
the a `r Biocpkg("DNAcopy")` object which will be quite helpful for plot
functions.

## File formats

The package can handle different file formats: GRanges, ExpressionSet objects,
matrix or a vector.

### Vector

```{r}
data2 <- data[, 1]
res <- fastseg(data2)
head(res)
```

### Matrix

```{r}
data2 <- data[1:400, ]
res <- fastseg(data2)
head(res)
```

### GRanges objects

```{r}
library("GenomicRanges")

## with both individuals
gr <- GRanges(seqnames=chrom,
        ranges=IRanges(maploc, end=maploc))
mcols(gr) <- data
colnames(mcols(gr)) <- samplenames
res <- fastseg(gr)
head(res)

## with one individual
gr2 <- gr
data2 <- as.matrix(data[, 1])
colnames(data2) <- "sample1"
mcols(gr2) <- data2
res <- fastseg(gr2)
head(res)
```

### ExpressionSet objects

```{r message=FALSE}
library(oligo)
eSet <- new("ExpressionSet")
assayData(eSet) <- list(intensity=data)

featureData(eSet) <- new("AnnotatedDataFrame", 
        data=data.frame(
                chrom = paste("chr",chrom,sep=""),
                start = maploc, 
                end   = maploc,stringsAsFactors=FALSE))
phenoData(eSet) <- new("AnnotatedDataFrame", 
        data=data.frame(samples=samplenames))
sampleNames(eSet) <- samplenames
res <- fastseg(eSet)
head(res)
```

### Vector

```{r}
data2 <- data[, 1]
res <- fastseg(data2)
head(res)
```

### Matrix

```{r}
data2 <- data[1:400, ]
res <- fastseg(data2)
head(res)
```

## Plotting the segmentation results

For plotting the data we have to generate an `r Biocpkg("DNAcopy")` object out
of the segmentation results:

```{r}
## with both individuals
gr <- GRanges(seqnames=chrom,
        ranges=IRanges(maploc, end=maploc))
mcols(gr) <- data
colnames(mcols(gr)) <- samplenames
res <- fastseg(gr,segMedianT=0.2)
```

The plotting is done via the `plot` function of `r Biocpkg("DNAcopy")`:

```{r fig=TRUE}
segPlot(gr,res, plot.type="w")
```

Or alternatively:

```{r fig.height=10, fig.width=10, message=FALSE, fig=TRUE}
segPlot(gr, res, plot.type="s")
```

## Performance of the method

Here we show that `r Biocpkg("fastseg")` outperforms `r Biocpkg("DNAcopy")` with
respect to computational time on summarized microarray data. The quality of the
segmentation result of both `r Biocpkg("fastseg")` and `r Biocpkg("DNAcopy")`
depends strongly on the methods' parameters.\ The data is a small subset of copy
number calls which were produced by the `cn.farms` algorithm @Clevert:11 from
an Affymetrix SNP microarray experiment of a HapMap sample.

```{r}
data(fastsegData)
system.time(res <- fastseg(fastsegData))
```

```{r echo=TRUE, fig=TRUE}
segPlot(fastsegData,res, plot.type="w")
```

```{r message=FALSE}
library(DNAcopy)
cna <- DNAcopy::CNA(fastsegData,chrom="chr1",maploc=1:length(fastsegData))
system.time(res2 <- DNAcopy::segment(cna))
```

```{r echo=TRUE, fig=TRUE}
plot(res2, plot.type="w", xmaploc=TRUE)
```

# Future Extensions

We are planning to program a parallelized version of this package.
Furthermore we will enhance the plot functions by our own.

# How to cite this package

If you use this package for research that is published later, you are kindly
asked to cite it as follows: [@Klambauer:12].

To obtain BibTeX entries of the two references, you can enter the following into
your R session:

```{r eval=FALSE}
toBibtex(citation("fastseg"))
```

[^1]: <http://www.nature.com/ng/journal/v29/n3/suppinfo/ng754_S1.html>
